---
title: "Metabolite-Cytokine Correlation Plot"
output: html_notebook
---

## Introduction

This R Markdown document analyzes repeated-measures correlation between cytokine and metabolomics data from the Rapid Administration of L-Carnitine in Septic Shock (RACE) clinical trial. Serum samples were collected at baseline (T0), 24-hours (T24), and 48-hours (T48) post-enrollment, and metabolic data was generated from LC-MS/MS (acylcarnitines) and 1H-NMR. This objective of this analysis is to calculate the repeated-measure correlation using the rmcorr package and investiagte the relationship between cytokines and the metabolic host-reponse in septic shock (i.e. saline placebo-treated patients). 

```{r}
library(ggcorrplot)
library(rmcorr)
```


```{r}
#prepare repeated measure dataframe 
#use the transformed dataframe (ac_p_trans, nmr_p_trans, and model_cyto_df from the LMM modeling notebook)

rmcorr_df <- ac_p_trans %>% 
  select(studyid, Timepoint, C10: L.carnitine) %>% 
  left_join(nmr_p_trans %>% 
             select(studyid, Timepoint, X2.Hydroxybutyrate:Valine),
            by =c('studyid', 'Timepoint')) %>% 
  left_join(model_cyto_df %>% 
              select(studyid, Fractalkine:ANG2, Timepoint),
                     by = c('studyid', 'Timepoint')) %>% 
  mutate(studyid = as.factor(studyid)) %>%
  na.omit() %>% 
  rename('2-Hydroxybutyrate' = X2.Hydroxybutyrate, 
         '2-Oxoisocaproate' = X2.Oxoisocaproate, 
         '3-Hydroxybutyrate' = X3.Hydroxybutyrate,
         'Propylene Glycol' = Propylene.glycol, 
         'Histidine' = Histidine.x)

#determine coloumns names for looping  
cols <- colnames(rmcorr_df[,3:63])

#testing the rmcorr function
rmcorr(participant = 'studyid', measure1 = 'C2', measure2 = 'C10', 
       dataset = rmcorr_df) 

#will give warning message when an analyte is correlated with itself
rmcorr(participant = 'studyid', measure1 = 'C2', measure2 = 'C2', 
       dataset = rmcorr_df) 
```


```{r}
loop_rmcorr_r <- function(df, var1, var2){
  fit <- rmcorr(participant = 'studyid', 
         measure1 = get(var1), 
         measure2 = get(var2),
         dataset = df)

  
  r <- as.data.frame(row.names = var2, fit$r)
  colnames(r) <- paste(var1)
  
  return(r)
}

loop_rmcorr_p <- function(df, var1, var2){
  fit <- rmcorr(participant = 'studyid', 
         measure1 = get(var1), 
         measure2 = get(var2),
         dataset = df)

  r <- as.data.frame(row.names = var2, fit$p)
  colnames(r) <- paste(var1)
  
  return(r)
}
```


```{r message=F, warning=F}
#message and warnings turned off because the loop will give warning message everytime an analyte is correlated with itself

#we remove these r=1 correlations when plotting 

r_df <- as_tibble()
for (y in cols) {
  r_df <- bind_rows(bind_cols(lapply(cols,
              function(x)
                loop_rmcorr_r(
                  var1 = x,
                  var2 = y,
                  df = rmcorr_df))),r_df)
}

p_df <- as_tibble()
for (y in cols) {
  p_df <- bind_rows(bind_cols(lapply(cols,
              function(x)
                loop_rmcorr_p(
                  var1 = x,
                  var2 = y,
                  df = rmcorr_df))),p_df)
}
```



```{r}
ggcorrplot(as.matrix(r_df %>% 
                       select(52:61)%>% 
                       filter(!row_number() %in% c(1:10))), 
           p.mat = as.matrix(p_df %>% 
                                select(52:61) %>%
                      filter(!row_number() %in% c(1:10))),
          insig = "blank", 
          outline.color	= 'black',
          ggtheme = theme_pubr(), 
          tl.cex = 10  ) + 
  scale_fill_gradient2(limit = c(-1, 1), 
                       low = "#235070" ,
                       high =  "#6c1d0e", 
                       mid = "white", 
                       midpoint = 0)
ggsave(filename = 'Figures/Figure3.tiff', width = 15, height = 15)



```

```{r}
ggcorrplot(as.matrix(r_df), 
           p.mat = as.matrix(p_df ),
          insig = "blank", 
          outline.color	= 'black',
          ggtheme = theme_pubr(), 
          tl.cex = 10
          ) + 
  scale_fill_gradient2(limit = c(-1, 1), 
                       low = "#235070" ,
                       high =  "#6c1d0e", 
                       mid = "white", 
                       midpoint = 0)
```

