---
title: "LMM RACE Placebo"
author: "Teddy Jennaro"
date: "February 9, 2023"
output: 
  html_document:
     df_print: paged
      toc: true
      toc_depth: 4
      toc_float:
        collapsed: no
        smooth_scroll: yes
---

```{r PackageLoad, echo=FALSE, message=FALSE, warning=FALSE}
library('ggpubr')
library('tidyverse')
library('knitr')
library('lme4')
library("sjPlot")
library('lattice')
library('ggeffects')
library('MetBrewer')
library('pbkrtest')
library('mixedup')
library('formulaic')
library('flextable')

```

## Introduction

This R Markdown document analyzes longitudinal metabolomics data from the Rapid Administration of L-Carnitine in Septic Shock (RACE) clinical trial. Serum samples were collected at baseline (T0), 24-hours (T24), and 48-hours (T48) post-enrollment, and metabolic data was generated from LC-MS/MS (acylcarnitines) and 1H-NMR. This objective of this analysis is to characterize the metabolic host-reponse in septic shock (i.e. saline placebo-treated patients). 

## Load and Process Data

```{r FileLoad, echo=FALSE}

#load in master CSV of RACE Data
file = 'Data/RACE_combined.csv'

df_master = read.csv(file)
```

```{r CodeVariables, echo=FALSE}
#make time numeric, create a 'day' variable, and create 28-day mortality
df_master %>% mutate(Time = recode(Timepoint, 
                           `T0`=0,
                           `T24`=24,
                           `T48`=48,
                           `T72`=72),
                     Day = Time / 24,
                     Mortality28 = case_when(
                       survivaldays <= 28 ~ 1,
                       survivaldays >28 ~0)) -> df_master

#consider only data from 0-48 hours in patients with:
#1) acylcarnitines measured; 2) treated with placebo (dose == 0);
ac_p = df_master %>% 
  filter(ACstatus==1 & Timepoint != 'T72' & dose == 0 )

```

```{r CheckCounts, echo=FALSE}
ac_p %>% group_by(Timepoint) %>% 
  summarize(n=n())

ac_p %>% distinct(studyid, .keep_all = TRUE) %>%
  group_by(Mortality28) %>% 
  summarize(n=n())

```


```{r PrepareAC}
#cols 39:62 represent the measurement of 24 acylcarnitine species
#importantly there is no missing data in this LC-MS/MS data. 
#First log transform the data, then standardize to a mean of zero and sd of 1. 
ac_p %>% 
  mutate_at(29:52, log) %>%
  mutate_at(29:52, scale) -> ac_p_trans
```

NMR Metabolomics has missing data. In metabolites with sufficient data (i.e. <30% missingness), we will compute missing values as the minimum concnetration observed divided by 2. Since serum samples were randomized into batches upon processing, we will impute missing values in the entire cohort (placebo & treated patients) and then filter out patients who received study drug. 
```{r PrepareNMR}
#first consider only data from 0-48 hours in patients with:
#1) NMR measured
nmr = df_master %>% 
  filter(NMRstatus==1)

#calculate the missing rate for each metabolite
#NMR metabolites represent columns 64:90
metab_missing_rate = apply(nmr[,54:80], MARGIN=2, FUN=function(x) { sum(is.na(x)) / length(x) } )

```

Since there are no metabolites with a missing rate > 30%, all measured metabolite columns can be considered further. 

```{r}
#impute missing as min observed / 2
nmr_metabolites = nmr[,54:80]
nmr_metabolites_imp_val = apply(nmr_metabolites, FUN=function(x) { min(x,na.rm = T) }, MARGIN = 2) / 2

#loop through & replace missing values 
for(j in 1:dim(nmr_metabolites)[2]) {
  imp_val = as.numeric(nmr_metabolites_imp_val[j])
  vals = as.numeric(nmr_metabolites[,j])
  nmr_metabolites[is.na(vals),j] = imp_val
}

#replace columns 
nmr_metabolites_matrix = as.matrix(nmr_metabolites)
nmr[,54:80] = data.frame(nmr_metabolites_matrix)
```


```{r}
nmr_p = nmr %>% 
  filter(dose == 0 & Timepoint != 'T72')

nmr_p %>% group_by(Timepoint) %>% 
  summarize(n=n())

nmr_p %>% distinct(studyid, .keep_all = TRUE) %>%
  group_by(Mortality28) %>% 
  summarize(n=n())
```

```{r}
nmr_p %>% 
  mutate_at(54:80, log) %>%
  mutate_at(54:80, scale) -> nmr_p_trans
```



```{r GetMetaboliteNames}
#this gets the column names from AC and NMR datasets
#we will use these later to loop in mixed model functions
ac_metabs = colnames(ac_p_trans)[29:52]
nmr_metabs = colnames(nmr_p_trans)[54:80]

```

```{r PrepareCytokineData}
#this will load in the cytokine data and prepare it for LMM
cyto_df <- read_csv('Data/RACE_CytokineData.csv')
model_cyto_df <- cyto_df %>% 
  mutate(Time = recode(Timepoint, 
                       `T0`=0,
                       `T24`=24,
                       `T48`=48,
                       `T72`=72),
         Day = Time / 24, 
         Sex = `Gender:`,
         Age_cent = Age - 59.63889, 
         T0_SOFA_cent = `Calculated T0 SOFA Value` - 11) %>% 
  mutate(across(c(Fractalkine:ANG2),log)) %>% 
  mutate(across(c(Fractalkine:ANG2), scale))

cyto_names <- model_cyto_df %>% 
  select(Fractalkine:ANG2) %>% 
  colnames()

cyto_names <- add.backtick(cyto_names)

```


```{r LMM_FXN}
#this is the function which will complete linear mixed modeling
#it selects fixed vs. interaction models based on the BIC
#it then calculates the overall p value in comparison to null model
#the user specifies the type of models to consider 
LMM <- function(metabs, eq1, eq2, eq3, df){
  print(metabs)
  null_eq = paste(metabs, eq1)
  null_mod = lmer(null_eq, data = df)
  #summary(null_mod)
  
  fixed_eq = paste(metabs, eq2)
  fixed_mod = lmer(fixed_eq, data = df)
  fixed_BIC = BIC(fixed_mod)
  fixed_summary = extract_fixed_effects(fixed_mod)
  #summary(fixed_mod)
  
  int_eq = paste(metabs, eq3)
  int_mod = lmer(int_eq, data = df)
  int_BIC = BIC(int_mod)
  int_summary = extract_fixed_effects(int_mod)
  #summary(fixed_mod)
  
  LRT1_2 = KRmodcomp(null_mod, fixed_mod)
  fixed_p = LRT1_2$stats$p.value
  
  LRT1_3 = KRmodcomp(null_mod, int_mod)
  int_p = LRT1_3$stats$p.value
  
  LRT2_3 = KRmodcomp(fixed_mod, int_mod)
  test_p = LRT2_3$stats$p.value
  
  if (test_p < 0.05) {
    overall_p = int_p
    
    overall_mod = "Interaction Model"
    
    Day = int_summary %>% 
      filter(term == 'Day') %>% 
      select(value, lower_2.5, upper_97.5) %>%  
      unite(int, lower_2.5:upper_97.5, sep = ', ') %>% 
      mutate(int = paste('(', int, ')', sep='')) %>% 
      unite(DaySlope, value:int, sep = " ") %>% pull
    
    Mortality28 = int_summary %>% 
      filter(term == 'Mortality28') %>% 
      select(value, lower_2.5, upper_97.5) %>%  
      unite(int, lower_2.5:upper_97.5, sep = ', ') %>% 
      mutate(int = paste('(', int, ')', sep='')) %>% 
      unite(DaySlope, value:int, sep = " ") %>% pull
    
    Int = int_summary %>% 
      filter(term == 'Day:Mortality28') %>% 
      select(value, lower_2.5, upper_97.5) %>%  
      unite(int, lower_2.5:upper_97.5, sep = ', ') %>% 
      mutate(int = paste('(', int, ')', sep='')) %>% 
      unite(DaySlope, value:int, sep = " ") %>% pull
    
    
  } else{
    overall_p = fixed_p
    
    overall_mod = "Fixed Effects Only Model"
    
    Day = fixed_summary %>% 
      filter(term == 'Day') %>% 
      select(value, lower_2.5, upper_97.5) %>%  
      unite(int, lower_2.5:upper_97.5, sep = ', ') %>% 
      mutate(int = paste('(', int, ')', sep='')) %>% 
      unite(DaySlope, value:int, sep = " ") %>% pull
        
    Mortality28 = fixed_summary %>% 
      filter(term == 'Mortality28') %>% 
      select(value, lower_2.5, upper_97.5) %>%  
      unite(int, lower_2.5:upper_97.5, sep = ', ') %>% 
      mutate(int = paste('(', int, ')', sep='')) %>% 
      unite(DaySlope, value:int, sep = " ") %>% pull
    
    Int = NA
  }
  
  return(c(metabs, overall_mod, overall_p, Day, Mortality28, Int))

  }



```

```{r}
#center co-variables 
nmr_p_trans$Age_cent <- nmr_p_trans$Age - 59.63889
nmr_p_trans$T0_SOFA_cent <- nmr_p_trans$Calculated.T0.SOFA.Value - 11

ac_p_trans$Age_cent <- ac_p_trans$Age - 59.63889
ac_p_trans$T0_SOFA_cent <- ac_p_trans$Calculated.T0.SOFA.Value - 11
```


```{r}
#Define the mixed models
#each conctains a patient level intercept
#null also has covariables
#fixed adds Day + Mortality 
#int adds Day * Mortality 
null = " ~ Age_cent + Sex + T0_SOFA_cent  +  (1 | studyid)"
fixed = "~ Age_cent + Sex + T0_SOFA_cent + Day + Mortality28 + (1 | studyid)"
int = "~  Age_cent + Sex + T0_SOFA_cent + Day * Mortality28 + (1 | studyid)"


```



```{r}
#run on NMR
nmr_LMM_results_RI = lapply(
  nmr_metabs, function(x) 
  LMM(metabs = x, 
    eq1 = null,
    eq2 = fixed, 
    eq3 = int, 
    df = nmr_p_trans))

#run on ACs
ac_LMM_results_RI = lapply(
  ac_metabs, function(x) 
  LMM(metabs = x, 
    eq1 = null,
    eq2 = fixed, 
    eq3 = int, 
    df = ac_p_trans))

#run on cytokines
cyto_res_LMM_results_RI = lapply(
  cyto_names, function(x) 
  LMM(metabs = x, 
    eq1 = null,
    eq2 = fixed, 
    eq3 = int, 
    df = model_cyto_df))

#bind together
#make p-value numeric; adjust according to FDR;
#rank-order
#rename
data.frame(do.call(rbind, nmr_LMM_results_RI) %>% 
             as_tibble() %>% 
             mutate(Platform = 'NMR')) %>% 
  bind_rows(data.frame(do.call(rbind, ac_LMM_results_RI) %>% 
                         as_tibble() %>%
                         mutate(Platform = 'LC-MS'))) %>% 
  bind_rows(data.frame(do.call(rbind, cyto_res_LMM_results_RI) %>% 
                         as_tibble() %>%
                         mutate(Platform = 'Protein Immunoassay'))) %>%
  mutate(p = as.numeric(V3)) %>% 
  mutate(`Q-Value` = p.adjust(p, method = 'fdr')) %>% 
  arrange(Platform, `Q-Value`) %>% 
  rename(Metabolite = V1,
         `Model Selected` = V2, 
         Time = V4, 
         Mortality = V5, 
         Interaction = V6) %>% 
  select(Metabolite, Platform, `Model Selected`, 
         `Q-Value`,
         Time, Mortality, Interaction) %>% 
  flextable() %>% 
  set_formatter(
        `Q-Value` = function(x) sprintf("%.02e", x)
      ) %>% 
  align(align ='center', part = 'all') %>% 
  align(j = 1, align ='left', part = 'all') %>% 

  bold(bold = TRUE, part = 'header') %>%
  flextable::autofit() -> ft2

tf <- tempfile(fileext = ".docx")
save_as_docx(ft2, path = tf)
         
```

```{r}
plotserieslines <- function(df, yvar, group){
  ggplot(data = df, aes_(x= ~Timepoint, y = as.name(yvar),
             group = ~Mortality28, color= ~Mortality28
            )) + 
    stat_summary(fun.data = "median_hilow", 
               fun.args=(conf.int=0.5), geom = 'errorbar',
               position = pd, width = 0.2) + 
    stat_summary(fun = 'median', geom = 'line', 
               mapping = aes(group= Mortality28),
               position = pd) + 
    stat_summary(fun = 'median', geom = 'point', 
               mapping = aes(group= Mortality28),
               position = pd) +
    stat_compare_means(label = 'p.signif', show.legend = F,
                       symnum.args = list(
                         cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                         symbols = c("****", "***", "**", "*", ""))) +
    scale_color_manual(name = "28-Day Status",
                       labels = c("Survivor", "Non-Survivor"),
                       values=met.brewer("Troy", 2)) + 
    geom_jitter(position = pd, alpha = 0.05)
}
```


```{r}
ac_p_plot <- ac_p %>% 
  mutate_at(.vars = 29:52, ~log(., base=10)) %>% 
  mutate(Mortality28 = as.factor(Mortality28))

nmr_p_plot <- nmr_p %>% 
  mutate_at(.vars = 54:80, ~log(., base=10)) %>% 
  mutate(Mortality28 = as.factor(Mortality28))

cyto_p_plot <- cyto_df %>%
  mutate_at(.vars = 8:17, ~log(., base=10)) %>% 
  mutate(Mortality28 = as.factor(Mortality28))
  
  
```

```{r}
#setting themes for plotting
theme_set(
  theme_linedraw(base_size = 13) +
    theme(
          panel.grid = element_blank(),
          strip.text = element_text(margin = margin(b = 1, t = 1)))
)
#dodge for space between outcomes on plots
pd <- position_dodge(width = 0.5)
```


```{r}
C2_p <- plotserieslines(df = ac_p_plot, yvar = 'C2') + 
  ylab (label = expression(Log[10] ~ paste(mu,M))) +
  ggtitle("Acetylcarnitine (C2)") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5)) 

C6_p <- plotserieslines(df = ac_p_plot, yvar = 'C6') + 
  ylab (label = expression(Log[10] ~ paste(mu,M))) +
  ggtitle("Hexanoylcarnitine (C6)") + 
  rremove('ylab') +
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5))

C8_p <- plotserieslines(df = ac_p_plot, yvar = 'C16') + 
  ylab (label = expression(Log[10] ~ paste(mu,M))) +
  ggtitle("Palmitoylcarnitine (C16)") + 
  rremove('ylab') +
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5))

ac_top_plots <- ggarrange(C2_p , C6_p, C8_p, nrow = 1,
          common.legend = TRUE, legend = 'none')
  

```

```{r}
Isoleucine_p <- plotserieslines(df = nmr_p_plot, yvar = 'Isoleucine') + 
  ylab (label = expression(Log[10] ~ paste(mu,M))) +
  ggtitle("Isoleucine") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5)) 

Lactate_p <- plotserieslines(df = nmr_p_plot, yvar = 'Lactate') + 
  ylab (label = expression(Log[10] ~ paste(mu,M))) +
  ggtitle("Lactate") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5)) 

Pyruvate_p <- plotserieslines(df = nmr_p_plot, yvar = 'Pyruvate') + 
  ylab (label = expression(Log[10] ~ paste(mu,M))) +
  ggtitle("Pyruvate") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5)) 

nmr_top_plots <- ggarrange(Isoleucine_p , Lactate_p, Pyruvate_p, nrow = 1,
          common.legend = TRUE, legend = 'none')

```


```{r}

IL_6_p <- plotserieslines(df = cyto_p_plot, yvar = 'IL-6') + 
  ylab (label = expression(Log[10] ~ pg ~ mL^-1)) +
  ggtitle("IL-6") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5))

IL_8_p <- plotserieslines(df = cyto_p_plot, yvar = 'IL-8') + 
  ylab (label = expression(Log[10] ~ pg ~ mL^-1)) +
  ggtitle("IL-8") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5))

TNF_p <- plotserieslines(df = cyto_p_plot, yvar = 'TNFα') + 
  ylab (label = expression(Log[10] ~ pg ~ mL^-1)) +
  ggtitle("TNFα") + 
  #theme_pubr() + 
  theme(plot.title = element_text(hjust = 0.5))

cyto_top_plots <- ggarrange(IL_6_p , IL_8_p, TNF_p, nrow = 1,
          common.legend = TRUE, legend = 'bottom')
```

```{r}
fig2 <- ggarrange(ac_top_plots, nmr_top_plots, cyto_top_plots,
          labels = c('A','B','C'),
          nrow =3, common.legend = TRUE)

ggsave(plot = fig2, 
       filename = 'Figures/Fig2.tiff',  
       width = 8.5, height = 13)
```

